---
title: "20230407_HALE_in_AML_proteomics"
author: "Sebastian Wolf"
toc: true
format:
  html:
    theme: litera
    self-contained: true
    code-fold: true
knitr:
  opts_chunk:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(MultiAssayExperiment)

prot_data <- readRDS("~/Forschung/AG Oellerich/Datensätze/FFM_Discovery_Cohort/FFM_Proteomics/prot_ffm.multiassayexperiment.RDS")

prot_data <- assay(prot_data, "proteomics_unimputed")

hlae <- "P13747"

phen_data <- read_csv("~/Forschung/AG Oellerich/Datensätze/FFM_Discovery_Cohort/FFM_AML_Discovery_Cohort.csv")

theme_set(cowplot::theme_minimal_grid() + cowplot::panel_border())
theme_update(legend.position = "bottom")
```

General distribution of HLA-E expression in the data set

```{r}
prot_data %>% 
  t() %>% 
  as_tibble(rownames  = "ID") %>% 
  dplyr::select(ID, "P13747") %>% 
  right_join(phen_data, by=c("ID")) %>% 
  dplyr::rename("HLA_E" = `P13747`) -> phen_data
  
phen_data %>%   
  ggplot(aes(x=HLA_E)) +
  geom_density(fill = "grey") 
```

Right-skewed expression with a bimodal peak.

PCA plot colored by HLA-E expression

```{r}
prot_data %>% 
  t() %>% 
  pcaMethods::pca() %>% 
  pcaMethods::scores() %>% 
  as_tibble(rownames = "ID") %>% 
  left_join(phen_data, by="ID") %>% 
  ggplot(aes(x=PC1, y = PC2, col = HLA_E)) +
  geom_point()
```

Correlation plot HLA_E expression and first 10 PCs

```{r}
prot_data %>% 
  t() %>% 
  pcaMethods::pca(nPcs = 10) %>% 
  pcaMethods::scores() %>% 
  as_tibble(rownames = "ID") %>% 
  left_join(phen_data, by="ID") %>% 
  dplyr::select(contains("PC"), HLA_E) %>% 
  cor() %>% 
  corrplot::corrplot()
```

```{r}
prot_data %>% 
  t() %>% 
  pcaMethods::pca(nPcs = 10) %>% 
  pcaMethods::scores() %>% 
  as_tibble(rownames = "ID") %>% 
  left_join(phen_data, by="ID") %>% 
  dplyr::select(PC1, PC5, HLA_E) %>% 
  pivot_longer(cols = c(PC1, PC5), names_to = "PC", values_to = "PC_score") %>% 
  ggplot(aes(x=PC_score, y = HLA_E, col = HLA_E)) +
  geom_point() +
  stat_smooth(method = "lm") +
  facet_wrap(. ~ PC, scale = "free") +
  ggpubr::stat_cor()
```

Moderately weak correlations between PCs 1 and 5 and HLA-E expression.

HLA-E expression and clinical features

```{r}
phen_data %>% 
  mutate(HLA_E_dic = factor(HLA_E >= median(HLA_E, na.rm = T)),
         HLA_E_quant = factor(cut(HLA_E, 4))) -> phen_data

vars <- c("Sex", "Age_group", "Precursor", "FAB", "WHO", "Complex_Karyotype", "NPM1", "FLT3", "ELN2017")

tableone::CreateTableOne(data = phen_data, strata = "HLA_E_dic", vars = vars) %>% 
  tableone::kableone(format = "html", caption = "Clinical parameters stratified by HLA-E expression")
  
```

```{r}
tableone::CreateTableOne(data = phen_data, strata = "HLA_E_quant", vars = vars) %>% 
  tableone::kableone(format = "html", caption = "Clinical parameters stratified by HLA-E expression")
```

HLA-E expression doesn't stratify along any clinically relevant axes.

```{r}
library(survival)
library(survminer)

ggsurvplot(survfit(Surv(OSM, OS_Status) ~ HLA_E_dic, phen_data), pval = T, risk.table = T, palette = "nejm")
```

```{r}
ggsurvplot(survfit(Surv(OSM, OS_Status) ~ HLA_E_quant, phen_data), pval = T, risk.table = T, palette = "nejm")
```

```{r}
ggsurvplot(survfit(Surv(EFSM, EFS_Status) ~ HLA_E_dic, phen_data), pval = T, risk.table = T, palette = "nejm")
```

HLA-E expression does not stratify long term outcome in our AML cohort.

# Summary

From the data set it seems that HLA-E expression in the data set is homogeneous and has no correlation with the clinical parameters tested. Furthermore it does not seem to predict clinical long term outcomes.

# Session info

last rendered `r Sys.Date()`

```{r}
sessionInfo()
```
